Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-01-31 draft.tex
.

\documentclass[10pt,a4paper]{article}
%\usepackage{layouts}
\usepackage[a4paper,margin=1.5in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{caption}
\captionsetup[table]{singlelinecheck=false}
\captionsetup[figure]{singlelinecheck=false}
\author{Fabian Schubert}

@@

Theoretical models of homeostasis are generally based on a form of feedback control. In control theory, this is also known as a \textit{closed loop control}. All these systems include a number of necessary functional compartments, which will be briefly introduced in the following. Concepts and terms were taken from \cite[p. 59-64]{Cybernetik_Systems_Cruse_2006}.
\newpage
\subsubsection{Theory of Control Systems} \label{Control_Sys_Theo_Section}
Generally, a closed loop control must contain some \emph{process} that is to be regulated. Despite its potential internal complexity, its functional role in the context of a feedback system is to simply generate a time-dependent output based on the history of a time-dependent input. This output is what one wishes to control and regulate by means of the feedback loop and thus represents the signal that is further processed. However, it might not be possible to directly quantify this desired output in terms of some internal state of the process. For example, the firing rate of a neuron is an abstract quantity, which can not be \emph{directly} related to some neuronal state variable. Consequently, it might be necessary to transform the actual output of the process into a different signal by a \textit{feedback transducer}. As a necessary next step, the (potentially transformed) output is fed into a \textit{comparator}, which, unsurprisingly, compares the given output of the process to some given target or \textit{reference}, typically by subtraction. Note that this reference can but does note have to be a time-dependent signal. However, the term \textit{homeostasis} generally refers to a fixed reference, while implementations with a time dependent reference can be regarded as a form of \textit{servomechanism}. The difference between reference and the feedback transducer's signal is known as the \textit{error} signal. This error signal is given to a \textit{controller}, which transforms this error signal into an appropriate input signal for the initially introduced process. Additionally, this is also the point in the control loop, where \textit{external disturbances} come into play by being added to the input generated by the controller. We illustrated the described flow of feedback control in Figure \ref{Closed_Loop_Control_Illustration}.
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/illustrations/closed_loop_control.eps}
\end{center}
\caption{Schematic illustration of a closed loop control as described in Section \ref{Control_Sys_Theo_Section}}
\label{Closed_Loop_Control_Illustration}
\end{figure}
Based on the way the controller generates input for the dynamic process based on the error signal, one can classify control loops by three types:
\newpage
\begin{itemize}
\item A \textit{proportional controller} generates a signal that is proportional to the error signal it receives. The most important property of a proportional controller is that the stationary state of the control loop generally does not imply an error signal equal to zero.
\item An \textit{integrator controller} uses an integration over the past of the error signal to obtain its output. The stationary solution necessarily implies that the error signal is zero.
\item A \textit{differential controller} generates its output by means of the time derivative of the error signal. This type of controller can quickly react to momentary changes but does not compensate for sustained deviations of the error signal from zero.
\end{itemize}
It should be pointed out that these types of control mechanisms can be simultaneously used in a control system, e.g. through a linear combination.

\subsubsection{Error Signals and Controllers of Neuronal Homeostasis}
When analyzing or modeling a biological system, a key step characterizing a control loop should be to identify the previously described compartments. While the functional separation illustrated in Figure \ref{Closed_Loop_Control_Illustration} might seem intuitive and "straight forward", it is not obvious which physiological element should be attributed to a certain abstract element.
As a start, we state that the process and its output controlled by homeostasis is the generation of action potentials and the resulting firing rate. As stated earlier, "firing rate" is an abstract quantity, requiring a feedback transducer to translate the neuron's spiking activity into a "sensible" quantity. In a nerve cell, this could be represented by e.g. the relation between the firing of action potentials and the temporary rise in $\mathrm{Ca^{2+}}$ concentration. An example for a homeostatic model making use of intracellular calcium concentration as a measure of spiking activity is the LeMasson-Abbott model \cite{LeMasson_1993}. More complex chains of activity encoding are of course possible, and we will introduce a recent model making additional use of the concentration of nitric oxide in Section \ref{DIff_Hom_Intro_Section} and \ref{NO_Experiments_Section}. Given a representation of activity has been detected, its deviation from a reference value will cause other physiological compartments to affect the neuron's ability to generate action potentials.

  , they necessarily contain an error estimation. This could be based e.g. on the neuron's firing rate of the recent past measured in a given window, or, on a physiological level, e.g. a concentration of some molecule within some part of the cell structure. In any case, the error estimation requires setting some predefined target that the feedback control will attempt to achieve. This, of course, raises the question of what would be an appropriate choice of homeostatic targets.

Experimental studies have found that neuronal populations exhibit a strong heterogeneity within their individual activity. In particular, many statistical analyses found evidence for a log-normal like distribution of firing rates \cite{Buzsaki_Fir_Rates_2014,Wohrer_Fir_Rates_2012}. Following these results, one approach of modeling this feature would be to randomly distribute individual targets according to the aforementioned statistics. A problem of this procedure is the absence of any particular deeper physiological motivation or causation. In the light of this conceptual problem, one could, due to a lack of better knowledge, simply define a single global target of activity. However, it has been shown that heterogeneity of neural response onto external input within a population is crucial for efficient stimulus encoding \cite{Marsat_2010}. Allowing for a broad distribution of activity thus has to be regarded an important compartment of models of cortical activity and structure.

@@

\subsection{Thesis Outline}
Key aspects of this thesis include an analysis of the stability of the homeostatic control, followed by a comparison of features of the original LIF-SORN and the diffusive variant. We expect to observe a preservation of non-random features that have been found in the original LIF-SORN while incorporating a stronger variance within neural activity which has previously been suppressed by the rigid single-cell homeostasis. In the face of possible \emph{new} features within the network's structure, we will clarify the causal relation between diffusive spatial interaction and synaptic topology.

@@

\subsubsection{Neuron Model}

We used a leaky integrate-and-fire-model for all neurons in the network, whose dynamics are described by a stochastic differential equation:
\begin{equation}
{\tau_m}dV = -(V-E_l)dt + \sqrt{\tau_m} \sigma dW
\label{LIF_Dynamics}
\end{equation}
where $V$ is the membrane potential, $E_l$ is the equilibrium membrane potential, $\tau_m$ is the time constant of the membrane, $\sigma$ is the standard deviation of the noise term and $dW$ is the standard Wiener process. A neuron is said to spike when its membrane potential reaches the threshold voltage $V_t$. The voltage is then reset to $V_r$. A refractory period was not implemented. A presynaptic spike causes a simple (delayed, see Table \ref{syn_conn_params}) increment of the membrane potential of the postsynaptic neuron by $w^{effective}_{ji}$. Table \ref{LIF_neuron_params} summarizes the aforementioned set of parameters.
\begin{table}
\begin{tabular}{|l|l|l|}
\hline
\textbf{parameter} & \textbf{exc. neur.} & \textbf{inh. neur.}\\ \hline
$\mathrm{E_l}$ & $\mathrm{-60\;mV}$ & $\mathrm{-60\;mV}$ \\ \hline
$\mathrm{\tau_m}$ & $\mathrm{20\;ms}$ & $\mathrm{20\;ms}$ \\ \hline
$\mathrm{V_r}$ & $\mathrm{-70\;mV}$ & $\mathrm{-60\;mV}$ \\ \hline
$\mathrm{\sigma}$ & $\mathrm{\sqrt{5}\;mV}$ & $\mathrm{\sqrt{5}\;mV}$ \\ \hline
$\mathrm{V_t}$ & subject to IP & $\mathrm{-58\;mV}$ \\ 
\hline
\end{tabular}
\caption{Parameters of LIF neuron}
\label{LIF_neuron_params}
\end{table}

@@

Apart from dynamic processes within synapses which contribute to a stabilization of the network's activity, neurons possess internal mechanisms capable of maintaining a desired regime of activity. Regular-spiking cells are known to down-(up-)regulate their firing rate upon increased (decreased) input on a timescale of tens of milliseconds \cite{Connors_Gutnick_Spike_Patterns,Benda_Herz_Spike_Frequ_Adaption}. Since our simulation did not incorporate any rapidly changing external drive, the network itself was not expected to exhibit fast changes of synaptic input, allowing us to neglect this feature. On the other hand, a similar form of adaption as a reaction on deprived or enhanced input can be observed on a timescale of hours to days \cite{Desai_IP}. In contrast to the former short-term adaption, which can be explained by a separation of timescales among different ionic currents in the cell \cite[p.~252-256]{Izhikevich_Dynsys}, in the latter case, \cite{Desai_IP} finds evidence that a long-term change in excitability can be attributed to an altered resistance of ionic channels.

In the original LIF-SORN, a simple form of low intrinsic homeostasis was implemented by altering the neurons' firing threshold based on the deviation from a target firing rate. This thesis implemented a new model of slow intrinsic homeostasis, based on the work in \cite{Sweeney_Paper}. The following section describes both models in further detail.

\subsubsection{Modeling of Homeostatic Intrinsic Plasticity}

@@

\section{Results} \label{results}

@@

On a larger, rate-based level, dynamical models of interconnected excitatory and inhibitory groups can exhibit sustained oscillations of activity \cite[p.~270]{Theor_Neur_Dayan}. As a simplest model, one can describe the dynamics of the excitatory and inhibitory populations as a two-dimensional dynamical system, representing the respective population rates $r_e$ and $r_i$:
\begin{align}
\tau_e \dot{r}_e  &= -r_e + \phi_e(r_e W_{ee} + r_i W_{ie}) \label{2d-exc_inh_rate_1} \\
\tau_i \dot{r}_i  &= -r_i + \phi_i(r_e W_{ei} + r_i W_{ii}) \label{2d-exc_inh_rate_2}
\end{align}

>>>>

\documentclass[10pt,a4paper]{article}
%\usepackage{layouts}
\usepackage[a4paper,margin=1.5in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{caption}

\usepackage{float}

\captionsetup[table]{singlelinecheck=false}
\captionsetup[figure]{singlelinecheck=false}
\author{Fabian Schubert}

@@

Theoretical models of homeostasis are generally based on a form of feedback control. In control theory, this is also known as a \textit{closed loop control}. All these systems include a number of necessary functional compartments, which will be briefly introduced in the following. Concepts and terms were taken from \cite[p. 59-64]{Cybernetik_Systems_Cruse_2006}.
\newpage
\subsubsection{Theory of Control Systems} \label{Control_Sys_Theo_Section}
Generally, a closed loop control must contain some \emph{process} that is to be regulated. Despite its potential internal complexity, its functional role in the context of a feedback system is to simply generate a time-dependent output based on the history of a time-dependent input. This output is what one wishes to control and regulate by means of the feedback loop and thus represents the signal that is further processed. However, it might not be possible to directly quantify this desired output in terms of some internal state of the process. For example, the firing rate of a neuron is an abstract quantity, which can not be \emph{directly} related to some neuronal state variable. Consequently, it might be necessary to transform the actual output of the process into a different signal by a \textit{feedback transducer}. As a necessary next step, the (potentially transformed) output is fed into a \textit{comparator}, which, unsurprisingly, compares the given output of the process to some given target or \textit{reference}, typically by subtraction. Note that this reference can but does note have to be a time-dependent signal. However, the term \textit{homeostasis} generally refers to a fixed reference, while implementations with a time dependent reference can be regarded as a form of \textit{servomechanism}. The difference between reference and the feedback transducer's signal is known as the \textit{error} signal. This error signal is given to a \textit{controller}, which transforms this error signal into an appropriate input signal for the initially introduced process. Additionally, this is also the point in the control loop, where \textit{external disturbances} come into play by being added to the input generated by the controller. We illustrated the described flow of feedback control in Figure \ref{Closed_Loop_Control_Illustration}.
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/illustrations/closed_loop_control.eps}
\end{center}
\caption{Schematic illustration of a closed loop control as described in Section \ref{Control_Sys_Theo_Section}}
\label{Closed_Loop_Control_Illustration}
\end{figure}
Based on the way the controller generates input for the dynamic process based on the error signal, one can classify control loops by three types:
\newpage
\begin{itemize}
\item A \textit{proportional controller} generates a signal that is proportional to the error signal it receives. The most important property of a proportional controller is that the stationary state of the control loop generally does not imply an error signal equal to zero.
\item An \textit{integrator controller} uses an integration over the past of the error signal to obtain its output. The stationary solution necessarily implies that the error signal is zero.
\item A \textit{differential controller} generates its output by means of the time derivative of the error signal. This type of controller can quickly react to momentary changes but does not compensate for sustained deviations of the error signal from zero.
\end{itemize}
It should be pointed out that these types of control mechanisms can be simultaneously used in a control system, e.g. through a linear combination. Furthermore, the presented terms refer to concepts of linear control theory. Thus, in the study of nonlinear feedback systems they only apply in a limited regime where a first order approximation of the system is sufficiently exact.

\subsubsection{Error Signals and Controllers of Neuronal Homeostasis}
When analyzing or modeling a biological system, a key step characterizing a control loop should be to identify the previously described compartments. While the functional separation illustrated in Figure \ref{Closed_Loop_Control_Illustration} might seem intuitive and "straight forward", it is not obvious which physiological element should be attributed to a certain abstract element.
As a start, we state that the process and its output controlled by homeostasis is the generation of action potentials and the resulting firing rate. As stated earlier, "firing rate" is an abstract quantity, requiring a feedback transducer to translate the neuron's spiking activity into a "sensible" quantity. In a nerve cell, this could be represented by e.g. the relation between the firing of action potentials and the temporary rise in $\mathrm{Ca^{2+}}$ concentration. An example for a homeostatic model making use of intracellular calcium concentration as a measure of spiking activity is the LeMasson-Abbott model \cite{LeMasson_1993}. More complex chains of activity encoding are of course possible, and we will introduce a recent model making additional use of the concentration of nitric oxide in Section \ref{DIff_Hom_Intro_Section} and \ref{NO_Experiments_Section}. Given a representation of activity, its deviation from a reference value will cause other physiological compartments to affect the neuron's ability to generate action potentials. Referring to the concepts introduced in the previous section, one could then ask what type of control best resembles the actual neuronal control. The fact that neurons, whose firing rate was artificially increased or suppressed over a long time, can reliably tune their activity back to the initial state (see e.g. \cite{Burrone_2002}) rules out a purely differential type of control. The return to baseline under different disturbances seems to point towards an integral-like control. However, it is clear that a the ability of a biological system to encode the accumulation of an error signal is limited by the finite amount of available resources. For example, the conductance of an ion channel has natural upper and lower bounds that it can not surpass. We will pick up this issue in Section \ref{Possible_Modifications_Section}.

In any case, modeling homeostasis includes an error estimation, which requires setting some predefined target that the feedback control will attempt to achieve. This, of course, raises the question of how to choose these homeostatic targets. Experimental studies have found that neuronal populations exhibit a strong heterogeneity within their individual activity. In particular, many statistical analyses found evidence for a log-normal like distribution of firing rates \cite{Buzsaki_Fir_Rates_2014,Wohrer_Fir_Rates_2012}. Following these results, one approach of modeling this feature would be to randomly distribute individual targets according to the aforementioned statistics. A problem of this procedure is the absence of any particular deeper physiological motivation or causation. In the light of this conceptual problem, one could, due to a lack of better knowledge, simply define a single global target of activity. However, it has been shown that heterogeneity of neural response onto external input within a population is crucial for efficient stimulus encoding \cite{Marsat_2010}. Allowing for a broad distribution of activity thus has to be regarded an important compartment of models of cortical activity and structure.

@@

\subsection{Thesis Outline}
Key aspects of this thesis include an analysis of the stability of the homeostatic control, followed by a comparison of features of the original LIF-SORN and the diffusive variant. We expect to observe a preservation of non-random features that have been found in the original LIF-SORN while incorporating a stronger variance within neural activity which has previously been suppressed by single-cell homeostasis. By an analytic approach we predict activity within the network, thereby gaining an understanding of the relation between spatial structure and firing rates. In the face of possible new features within the network's structure, we clarify the causal relation between diffusive spatial interaction and synaptic topology.

@@

\newpage
\subsubsection{Neuron Model}
We used a leaky integrate-and-fire-model for all neurons in the network, whose dynamics are described by a stochastic differential equation:
\begin{equation}
{\tau_m}dV = -(V-E_l)dt + \sqrt{\tau_m} \sigma dW
\label{LIF_Dynamics}
\end{equation}
where $V$ is the membrane potential, $E_l$ is the equilibrium membrane potential, $\tau_m$ is the time constant of the membrane, $\sigma$ is the standard deviation of the noise term and $dW$ is the standard Wiener process. A neuron is said to spike when its membrane potential reaches the threshold voltage $V_t$. The voltage is then reset to $V_r$. A refractory period was not implemented. A presynaptic spike causes a simple (delayed, see Table \ref{syn_conn_params}) increment of the membrane potential of the postsynaptic neuron by $w^{effective}_{ji}$. Table \ref{LIF_neuron_params} summarizes the aforementioned set of parameters.
\begin{table}[H]
\begin{tabular}{|l|l|l|}
\hline
\textbf{parameter} & \textbf{exc. neur.} & \textbf{inh. neur.}\\ \hline
$\mathrm{E_l}$ & $\mathrm{-60\;mV}$ & $\mathrm{-60\;mV}$ \\ \hline
$\mathrm{\tau_m}$ & $\mathrm{20\;ms}$ & $\mathrm{20\;ms}$ \\ \hline
$\mathrm{V_r}$ & $\mathrm{-70\;mV}$ & $\mathrm{-60\;mV}$ \\ \hline
$\mathrm{\sigma}$ & $\mathrm{\sqrt{5}\;mV}$ & $\mathrm{\sqrt{5}\;mV}$ \\ \hline
$\mathrm{V_t}$ & subject to IP & $\mathrm{-58\;mV}$ \\ 
\hline
\end{tabular}
\caption{Parameters of LIF neuron}
\label{LIF_neuron_params}
\end{table}

@@

Apart from dynamic processes within synapses which contribute to a stabilization of the network's activity, neurons possess internal mechanisms capable of maintaining a desired regime of activity. Regular-spiking cells are known to down-(up-)regulate their firing rate upon increased (decreased) input on a timescale of tens of milliseconds \cite{Connors_Gutnick_Spike_Patterns,Benda_Herz_Spike_Frequ_Adaption}. Since our simulation did not incorporate any rapidly changing external drive, the network itself was not expected to exhibit fast changes of synaptic input, allowing us to neglect this feature. On the other hand, a similar form of adaption as a reaction on deprived or enhanced input can be observed on a timescale of hours to days \cite{Desai_IP}. In contrast to the former short-term adaption, which can be explained by a separation of timescales among different ionic currents in the cell \cite[p.~252-256]{Izhikevich_Dynsys}, in the latter case, \cite{Desai_IP} finds evidence that a long-term change in excitability can be attributed to an altered resistance of ionic channels. In the original LIF-SORN, a simple form of low intrinsic homeostasis was implemented by altering the neurons' firing threshold based on the deviation from a target firing rate. This thesis implemented a new model of slow intrinsic homeostasis, based on the work in \cite{Sweeney_Paper}. The following section describes both models in further detail.
\subsubsection{Modeling of Homeostatic Intrinsic Plasticity}

@@

\newpage
\section{Results} \label{results}

@@

On a larger, rate-based level, dynamical models of interconnected excitatory and inhibitory groups can exhibit sustained oscillations of activity \cite[p.~270]{Theor_Neur_Dayan}. As a simplest model, one can describe the dynamics of the excitatory and inhibitory populations as a two-dimensional dynamical system, representing the respective population rates $r_e$ and $r_i$:
\tau_e \dot{r}_e  &= -r_e + \phi_e(r_e W_{ee} + r_i W_{ie}) \label{2d-exc_inh_rate_1} \\
\tau_i \dot{r}_i  &= -r_i + \phi_i(r_e W_{ei} + r_i W_{ii}) \label{2d-exc_inh_rate_2}
